Refinement for open automata

### 2021-06-16 2021-Refinement/report.tex
2 comments

\section{Introduction}
The open automata model was originally designed to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
This model is based around a hierarchical composition of generalised automata.
Verification is the process of checking properties on a description in a specific model (for instance an automaton).
This properties can be for instance being in some type or being deadlock free (a deadlock is an unintended state where no further action is possible).
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify some properties of applications \cite{qin:01823507, ameurboulifa:01526055} based on BIP (Behaviour-Interaction-Priority) \cite{basu:hal-00722395} and GCM (Grid Component Model) \cite{baude:inria-00323919}.
However equivalence relations are not always sufficient to perform verification, refinement relations are also used.
A refinement is the verification that an implementation satisfies a specification, whereas an equivalence would check that their behaviours are the same.
For instance a usecase on open pNet where equivalence is not sufficient can be found in an article about algorithms fo weak bisimulation on open automata \cite{wang:03126313}.

The contributions of this report are:
\begin{enumerate}
\item Formalisation of composition for open automata without requiring open pNets.
\item The characterisation of a composition that doesn't introduce deadlocks.
\item The introduciton of several refinement relations for open automata that behave nicely with composition.
\end{enumerate}
% TODO: Related work: thesis

@@

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter can only be used if there is a generating expression\Ludo{c est vrai pour les exemples mais pas pour les defs generiques (OA, bisim, ...): reformuler}\Quentin{Je pense que c'est aussi vrai pour les defs génériques, on définit une famille à l'aide d'une expression génératrice, par exemple \(\mpar{\beta_j}_{j \in J}\) est en réalité \(j \mapsto \beta_j\); une fonction est une expression génératrice}; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
However \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no generating expression and is represented here with the finite version of first notation.

@@

Notations are parameterised by a term algebra \(\terms\).
As any term algebra it has constant symbols with arity, variables, and a typing mechanism to distinguish well-formed and ill-formed terms.
The action algebra \(\actions\) is another term algebra.
It can be \Ludo{''can be'' is not much informative}\Quentin{Intended use, it is not mandatory} a subset of the terms.
The formulas \(\formulas\) are at least the first order formulas over \(\terms\) and \(\actions\).
\end{defi}
The term algebras are arbitrary.
The formulas contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality but it is more expressive to consider that the two terms are equal).

@@

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise it doesn't invalidates theorems because it can stand as a shorthand for substitution with any term which has the kept value as an interpretation\Ludo{comprends pas la fin de phrase, anglais?}\Quentin{Il manquait des mots}.
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulas\)]
\defitem The satisfiability of a formula \(f \in \formulas\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulas\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \bigg((a \uplus b) \vdash \Big((x \wedge y) \implies \exists z, P\mpar{x,z}\Big)\bigg) \]
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.
A pair of a state and a valuation is called a configuration.

@@

The initial valuation is a partial valuation of the variables.
If an undefined variable is set before being read, it behaves like an initially set variable.
If an undefined variable is not set before being read then its value may be any fixed value.
When we will introduce refinements in Section \ref{sec:prelref}, setting an undefined variable will be considered a valid refinement, for instance a \(5\) bits register is a particular \(n\) bits register.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
\end{defi}
An open transition can have many unbound variables.
Actually an effective transition of the automaton is any well-formed substitution of the unbound variables of the transition minus the automaton variables.

@@

The FH-bisimulation is currently the only equivalence relation on open automata.
The definition is not important, but the interested reader can check the article cited.
\end{noti}
To illustrate these definitions we consider two implementations (figure \ref{fig:enable}) of the LOTOS \cite{ISOLOTOS} operator enable in the open automata model.
The enable operator runs the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\mpar{t}\) (with \(t\) some data) that is synchronised with the first action of the right hand side agent which must be \(\act{accept}\mpar{t}\) (here \(t\) is an input), then only the latter agent runs.
The \(\act{accept}\) action is shortened as \(\act{acc}\) in the examples.
During the synchronised action, the value returned by the first agent is passed to the second agent.
\begin{exi}[Enable, state-oriented]
Graphical convention for drawing automata are as follows.
As standard automata, circles represent states and simple arrows represent transitions.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.
The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the emmited action is on the bottom side of open transitions.
Initial valuations are indicated near a double lined arrow pointing to the initial state.

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right data oriented}
\label{fig:enable}
\end{figure}
The automaton on the left side of the figure is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{\forall y, x \neq \delta\mpar{y}}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that the transition in the middle could have been expressed as \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\).
It would have avoided many effective transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\), which has the guard \(1 = 2\). \Ludo{I believe that with the usual definition of the semantics of open automata it is the same ... please check: you do all potential transitions with the conditions that x is the same in both. In any case as the matching is not linear, the simplification is not obvious}\Quentin{Same behaviour as here with the standard definition: RR-9177, sec 5.2}

@@

We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transitions have been omitted for the sake of simplifying the first example of open automata.
\end{exi}
Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT{r}\) are called the out-transitions of \(r\).
\(\fIT{r}\) are called the in-transitions of \(r\).
The local variables of a transition are all variables appearing in that transition except the global variables of the automaton.
\begin{align*}
	\fOT{r} & \defnotation \mset{\OTg \in T}{s = r} &
	\fIT{r} & \defnotation \mset{\OTg \in T}{s' = r} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}
The goal of the extractor \(\fvars{t}\) when \(t\) is a transition is to get variables that are only present in that transition.
Also transition variables sets are disjoint from one transition to another.
Other variables like automaton variables are already known at this point so there is no use in getting their name again.
Also the use of this extractor benefits from this exclusion: otherwise there would always be \(\setminus V\) following it to prevent variable shadowing.

\Ludo{is it really variable shadowing?}\Quentin{I don't know if the term is the same in math}

\Ludo{you do not necessarily need to explain everything if this is what you need, but you miss an intuition of what this set is}\Quentin{I don't know what else I could say to give the intuition, all elements are here, in my opinion, the only thing left to build the intuition is an example}

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = L \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = L \wedge x = y}{\mbrc{v \gets R}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = R}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar\Ludo{makes no sense: FH bisimlar is still undefined, neither formally nor informally}\Quentin{I can remove this sentence, it adds nothing} to the previous one \cite{henrio:01299562}.
The variables that are said to be local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
When taking effective transitions, its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = L \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{L = L \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto L\)).

An run of this open automaton can be: The automaton in the hole \(l\) emits many actions, this synchronises with the only transition possible at that moment because \(v = L\).
Then the automaton in hole \(l\) emits a \(\delta\mpar{t}\), this is synchonised with the action that sets \(v \gets R\) and the automata in hole \(r\) must emit a \(\act{acc}\mpar{t}\).
Finally the automaton in the hole \(r\) emits any sequence of actions.

Another run can be: The automaton in the hole \(l\) emits a sequence of non \(\delta\mpar{t}\) actions so the automaton in the hole \(r\) cannot emit a \(\act{acc}\mpar{t}\) and never runs.
\end{exi}
From this point open automata and open transitions are sometimes called automata and transitions for simplicity. % NEED NORMALISATION, LEFT IN CASE OF OMISSION

@@

\section{Properties of a refinement relation for Open Automata}\label{sec:proofelts}
There are several properties we may expect from a refinement relation.
Depending on these properties, different kinds of refinement can be used.
For example if we are interested in producing the same sequences of actions as another automaton we may want to use trace set inclusion as a refinement.
In this section we will present the main expected properties, which are related to composition.
Because of composition, a relation as strong as simulation must be used.

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
Generally the second relation is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property on the relation on states, so only talking about \(R\) by defining \(P\) is sufficient.\Ludo{je comprends pas cette phrase}\Quentin{Est-ce mieux?}
However later on in this article we will define more complex relations on automata so we clarify the distinction now.

A relation of second kind will be called ``relation on automata" whereas a relation of the first kind is called ``relation on configurations" as defined below.
In the work on FH-bisimulation \cite{henrio:01055091} such a relation \(R\) was defined as triples in \(S_1 \times S_2 \times \formulas\) where for \(\mpar{s_1, s_2} \in S_1 \times S_2\) there is a unique \(f \in \formulas\) such that \(\mpar{s_1, s_2, f} \in R\).
This is classically the definition of a function from \(S_1 \times S_2\) to \(\formulas\) so we will define it as such.
\begin{defi}[Relation on configurations]
A relation on configurations of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}
In this section we give properties on refinement relations, both on the relations on configurations and on the relation on automata.
When relating two automata \(A_1 \mathrel{\mathcal{R}} A_2\) or their configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\), \(A_1\) will be called the implementation and \(A_2\) the specification.

>>>>

\section{Introduction}
The open automata model was originally designed to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
 Verification is the process of checking that an implementation satisfies a specification.
\Ludo{verification is more general than that (or you need to refine specification): checking typing or absence of deadlocks belong to verification}
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify some properties of applications based on BIP  and GCM \cite{qin:01823507, ameurboulifa:01526055} \Ludo{cite and expand acronym (x2)}.
However equivalence relations are not always sufficient to perform verification, refinement relations are also used. \Ludo{verification/refinement/equivalence needs to be refined and explained}
For instance a usecase where open pNet where equivalence is not sufficient can be found  in an article about algorithms fo weak bisimulation on open automata \cite{wang:03126313}.

\Ludo{choose: either 1) you do not refer to open pnets below and explain informally open automata here (why are they useful, what makes them different from automata), or 2) you want to refer to pnets in the following and you explain informally pnets, uatomata and the relation}

The contributions of this article are \Ludo{report?} the definition of composition for open automata without having to define open pNets, and the introduction of several refinement relations for open automata. % TODO: expand

% TODO: Related work: thesis

\Ludo{RAPPEL: pour ecrire une intro: 1) on ecrit les contribs, 2) on explique avant les termeis techniques pour que le lecteur comprenne les contribs, 3) quand on a quasi fini le reste on ecrit le plan. ON peut commencer 1 et 2 un peu tot meme si il faut rediger le tout assez tard: on peut pas ecrire proprement un article sans avoir en tete les contribs de celui ci}

@@

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter can only be used if there is a generating expression\Ludo{c est vrai pour les exemples mais pas pour les defs generiques (OA, bisim, ...): reformuler}; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
However \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no generating expression and is represented here with the finite version of first notation.

@@

Notations are parameterised by a term algebra \(\terms\).
As any term algebra it has constant symbols with arity, variables, and a typing mechanism to distinguish well-formed and ill-formed terms.
The action algebra \(\actions\) is another term algebra.
It can be \Ludo{''can be'' is not much informative} a subset of the terms.
The formulas \(\formulas\) are at least the first order formulas over \(\terms\) and \(\actions\).
\end{defi}
The term algebras are arbitrary.
The formulas contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality but it is more expressive to consider that the two terms are equal).

@@

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise it doesn't invalidates theorems because it can stand as a shorthand for substitution with any term which is interpreted the kept value \Ludo{comprends pas la fin de phrase, anglais?}.
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulas\)]
\defitem The satisfiability of a formula \(f \in \formulas\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulas\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \bigg((a \uplus b) \vdash \Big((x \wedge y) \implies \exists z, P\mpar{x,z}\Big)\bigg) \]
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.
A pair of a state and a valuation is called a configuration.

@@

The initial valuation is a partial valuation of the variables.
If an undefined variable is set before being read, it behaves like an initially set variable.
If an undefined variable is not set before being read then its value may be any fixed value.
When refining \Ludo{bizarre de parler de rafinement ici: phrase a deplacer ou introduire par: Later we will consider refinement of autoamata and ....} the automaton it can be set by the implementation, or it can be left unspecified as a parameter of the automaton (for instance a \(n\) bits register has \(n\) as parameter).
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
\end{defi}
An open transition can have many unbound variables.
Actually an effective transition of the automaton is any well-formed substitution of the unbound variables of the transition minus the automaton variables.

@@

\Ludo{here the reader needs either an intuition of the FH bisim or a definition in any case, he needs to know that he will get one or the other and when. For example we give below an intuitive definition of the FH bisim based on examples.}

\Ludo{ If you need the notation but the def is not important you can explicitly say that the readers should refer to ?? for the details of the def, but here I do not believe it is wise OR you could say FH bisim is useless and you do not introduce the notation but you should not use it afterwards (again not adapted I believe)}

The FH-bisimulation is currently the only equivalence relation on open automata.
\end{noti}
To illustrate these definitions we consider two implementations (figure \ref{fig:enable}) of the LOTOS \cite{ISOLOTOS} operator enable in the open automata model.
The enable operator runs the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\mpar{t}\) (with \(t\) some data) that is synchronised with the first action of the right hand side agent which must be \(\act{accept}\mpar{t}\) (here \(t\) is an input), then only the latter agent runs.
The \(\act{accept}\) action is shortened as \(\act{acc}\) in the examples.
During the synchronised action, the value returned by the first agent is passed to the second agent.
\begin{exi}[Enable, state-oriented]
Graphical convention for drawing automata are as follows.
As standard automata, circles represent states and simple arrows represent transitions.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.
The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the emmited action is on the bottom side of open transitions.
Initial valuations are indicated near a double lined arrow pointing to the initial state.

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right data oriented}
\label{fig:enable}
\end{figure}
The automaton on the left side of the figure is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{\forall y, x \neq \delta\mpar{y}}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that the transition in the middle could have been expressed as \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\).
It would have avoided many effective transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\), which has the guard \(1 = 2\). \Ludo{I believe that with the usual definition of the semantics of open automata it is the same ... please check: you do all potential transitions with the conditions that x is the same in both. In any case as the matching is not linear, the simplification is not obvious}

@@

We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transitions have been omitted for the sake of simplifying the first example of open automata.
\end{exi}
Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT{r}\) are called the out-transitions of \(r\).
\(\fIT{r}\) are called the in-transitions of \(r\).
The local variables of a transition are all variables appearing in that transition except the global variables of the automaton.
\begin{align*}
	\fOT{r} & \defnotation \mset{\OTg \in T}{s = r} &
	\fIT{r} & \defnotation \mset{\OTg \in T}{s' = r} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}

The goal of the extractor \(\fvars{t}\) when \(t\) is a transition is to get variables that are unique \Ludo{what do you mean by unique here?} to that transition.
Other variables like automaton variables are already known at this point so there is no use in getting them \Ludo{this is not clear because I do not know what variables you ``NEED'' to ``GET'' (need and get are strange and vague to me)}.
Also the use of this extractor benefits from this exclusion: otherwise there would always be \(\setminus V\) following it to prevent variable shadowing.

\Ludo{is it really variable shadowing?}

\Ludo{you do not necessarily need to explain everything if this is what you need, but you miss an intuition of what this set is}

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is \Ludo{FH-bisimilar}{makes no sense: FH bisimlar is still undefined, neither formally nor informally} to the previous one \cite{henrio:01299562}. 
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds \Ludo{}{the intersection is at the same time empty and non-empty, why not use a different value (or underline them) or just say that they are in fact different names though it is easier to read using r and l for both}. \Ludo{le probleme ici c est qu on a l impression que tu essayes a la fois d etre tres strict et formel et pas vraiment en meme temps, il faut choisir}
The variables that are said to be local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
\Ludo{In a run,}{je comprends ps cette phrase:}  when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

An  run of this open automaton can be: The automaton in the hole \(l\) emits many actions, this synchronises with the only transition possible at that moment because \(v = l\).
Then the automaton in hole \(l\) emits a \(\delta\mpar{t}\), this is synchonised with the action that sets \(v \gets r\) and the automata in hole \(r\) must emit a \(\act{acc}\mpar{t}\).
Finally the automaton in the hole \(r\) emits any sequence of actions.

Another  run can be: The automaton in the hole \(l\) emits a sequence of non \(\delta\mpar{t}\) actions so the automaton in the hole \(r\) cannot emit a \(\act{acc}\mpar{t}\) and never runs.
\end{exi}

From this point open automata and open transitions are sometimes called automata and transitions for simplicity. % NEED NORMALISATION, LEFT IN CASE OF OMISSION

@@

\section{Properties of a refinement relation for Open Automata}\label{sec:proofelts}
There are several properties we may expect from a refinement relation.
Depending on these properties, different kinds of refinement can be used.
For example if we are interested in producing the same sequences of actions as another automaton we may want to use trace set inclusion as a refinement.
Here the main expected properties are related to composition and action refinement. % LATER: explain action refinement here or remove if not tackled
Because of composition, a relation as strong as simulation must be used.

\Ludo{New}{je suggere un paragraphe du style}
In this section we present a set of  properties that can be desirable for a refinement relation. We reach a point where we state a property that guarantees that the refinement does not create additional deadlock. Unfortunately, in the general case, this property is contradictory with the other ones, this is why we restrict the general composition defined before so that we only consider a composition  to be valid if it does not create deadlocks. With this new condition on validity of composition we are able to define in the next sections a refinement relation where most of the properties of interest are satisfied.

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
\Ludo
{Generally the second is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property, and only talking about \(R\) and \(P\) is sufficient.}{je comprends pas cette phrase}
However later on in this article we will define more complex relations on automata so we clarify the distinction now.
A relation of second kind will be called ``relation on automata" whereas a relation of the first kind is called ``relation on configurations" as defined below.
In the work on FH-bisimulation \cite{henrio:01055091} such a relation \(R\) was defined as triples in \(S_1 \times S_2 \times \formulas\) where for \(\mpar{s_1, s_2} \in S_1 \times S_2\) there is a unique \(f \in \formulas\) such that \(\mpar{s_1, s_2, f} \in R\).
This is classically the definition of a function from \(S_1 \times S_2\) to \(\formulas\) so we will define it as such.
\begin{defi}[Relation on configurations]
A relation on configurations of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}
The important properties we consider for the relation on automata are:
\begin{defi} A relation on open automata \(\leq\) is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a\subst{b}{j} \leq a\);
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \cong b\subst{c}{j}\);
\defitem \textbf{context refining for composition} iff \(\forall a\, b\, c, a \leq b \implies a\subst{c}{j} \leq b\subst{c}{j}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\subst{a}{j} \leq c\subst{b}{j}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\subst{a}{j} \leq d\subst{b}{j}\);
\defitem \textbf{compatible with FH-bisimulation} iff \(\forall a\, b\, c, d, a \cong b \wedge c \cong d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Correctness w.r.t.\@ composition means that every composition is considered a refinement. \Ludo{problem}{ you have not yet said that your relation was a refinement!}
This property captures the expected behaviour of a refinement relation on a compositional structure like the open automata.
Completeness w.r.t.\@ composition means that a refinement corresponds to the left automaton being equivalent to some composition of the right automaton.
This property is useful if we want to characterise the kind of operations that can generate any refined automaton starting from the specification.
\Ludo{}{I do not understand this: what is the domain of the valuation?}
We won't have it because extending the domain of the initial valuation will be a refinement which can't be expressed with composition.
Yet a ``completeness w.r.t.\@ some set of operations" is still a strong and interesting property.
Context refinement is the refinement being compatible with composition of the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the conjunction of the two latter (assuming the relation is a preorder).
This property is required on compositional structures to be of any use in verification, proofs and for model checking.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.
Since FH-bisimulation is the pre-existing equivalence relation, we should not introduce an incompatible proof strategy.
